---
title: "Reference Materials"
author: "Laurie Baker, Elaine Ferguson, Will Harvey, and Rachel Steenson"
date: "August 2019"
output: 
  html_document:
    theme: yeti
    highlight: tango
---


```{r Loading libraries and data, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(dplyr) # library for data wrangling
library(ggplot2) # library for plotting
library(lubridate) # library for working with and formatting dates
library(leaflet) # library for creating interactive maps using leaflet
library(rgdal) # library for colors
library(RColorBrewer) # library for colors
library(kableExtra) # library for constructing tables
library(stats) # library of stats

raw_data <- read.csv("data/raw_data.csv", stringsAsFactors = FALSE)
region_shp <- readOGR("data/TZ_Region_2012", "TZ_Region_2012")

mara_shp <- region_shp[which(region_shp$Region_Nam=="Mara"),]

region_shp$id <- rownames(region_shp@data)
region_shp_df <- fortify(region_shp, region = "id")
region_shp_df <- left_join(region_shp_df, region_shp@data, by = "id")

mara_shp$id <- rownames(mara_shp@data)
mara_shp_df <- fortify(mara_shp, region = "id")
mara_shp_df <- left_join(mara_shp_df, mara_shp@data, by = "id")

summary_data <- raw_data %>%
  filter(species=="human") %>%
  group_by(month, date, sex) %>%
  summarise(mean_age = mean(age),
            mean_density = mean(density)) %>%
  arrange(-mean_age)

summary_data_means <- raw_data %>%
  group_by(species) %>%
  summarise(mean_age = round(mean(age), digits=1),
            sd_age = round(sd(age), digits=1))

movement <- raw_data %>%
  group_by(month, species, region) %>%
  select(region, month, x,y) %>%
  filter(species == "dog" & region=="Mara")

# Create a colour palette
col_palette <- c("#231D51", "#178B8B", "#63C963", "#FFE31D")

```

<style>
h1, h2, h3, h4 {font-weight: bold;}
.question {font-weight: bold; color: #4E23B8;}
hr {border: 2px solid black;}
a:link {color: white; text-decoration: none;}
a:visited {color: white; text-decoration: none;}
a:hover {color: #c9bde9; text-decoration: none;}
a:active {color: white; text-decoration: none;}
.panel-aims {color: white; background-color: #4E23B8; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
.panel-section {color: black; background-color: #c9bde9; border-top: 3px solid black; border-bottom: 3px solid black; padding-left: 15px; padding-right: 15px; } 
</style>

<!--- ---------------------------------------------------------------------- -->
<!--- Aims section -->

<br>
<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-aims"> <!--- Begin panel -->
<h2>Contents</h2>
Click to view a specific section of the document.  

<a href="#1"><h3> 1. Getting to know your data</h3></a>
<ul>
<li><a href="#1.1">Loading in packages</a></li>
<li><a href="#1.2">Loading in data</a></li>
<li><a href="#1.3">Exploring data</a></li>
</ul>
<a href="#2"><h3> 2. Data subsetting and summarising</h3></a>
<ul>
<li><a href="#2.1">The dplyr package</a></li>
<li><a href="#2.2">Subsetting</a></li>
<li><a href="#2.3">Summarising</a></li>
<li><a href="#2.4">Adding new variables</a></li>
<li><a href="#2.5">The ``%`` (pipe) function</a></li>
</ul>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<!--- ---------------------------------------------------------------------- -->
<!--- 1. Getting to know the data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h2 id="1">Getting to know the data</h2>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->


<!--- ---------------------------------------------------------------------- -->
<!--- 1.1 Loading in packages -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1.1">Loading in packages</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>At the very beginning of any R script, you should load in any packages you plan on using.<.li>
<li>Packages contain functions to complete a <em>very</em> broad range of processes. </li>
<li>To load in your packages, we can use the ``library()`` function:</li>

``` {r, eval=FALSE}
library(dplyr)
```
</ul>
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 1.2 Loading in data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1.2">Loading in data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>As with packages, you should load in any data you want to use at the beginning on your R script.</li>
<li>This allows you to easily see what data is being loaded, rather than having to search through your code.</li>
<li>There are multiple functions to load in your data - we will only be using a <em>comma-seperated values</em> spreadsheet, so we need the ``read.csv()`` function:

``` {r, eval=FALSE}
raw_data <- read.csv("data/raw_data.csv") 
```
</ul>
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 1.3 Exploring data -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="1.3">Exploring data</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>We have several functions to get to know the data: 
<br>

``` {r, echo=FALSE}
explore <- data.frame("function name"=c("``View()``", "``nrow()``", "``names()``", "``dim()``", 
                                   "``summary()``", "``head()``", "``tail()``", "``str()``",
                                   "``unique()``", "``table()``", "``?``", "``??``"),
                      "description"=c("view the dataframe as a spreadsheet",
                                      "report the number of rows in the data frame",
                                      "report the column name in the data frame",
                                      "report the dimensions of the data frame",
                                      "report the summary statistics (counts, min, median, mean, max)",
                                      "print the first 6 rows of the data",
                                      "print the last 6 rows of the data",
                                      "report the variable type in each column (e.g. Factor, num (number), int (integer))",
                                      "report the unique elements of a variable",
                                      "calculate the count number for variables in the column(s) specified",
                                      "open the helpfile for a function contained in a loaded package",
                                      "open the helpfile for a function contained in an package that isn't loaded"))
kable_styling(kable(explore, col.names=c("function name", "description")), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```

</li>
</ul>
<br><br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2. Subsetting and summarising -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h2 id="2">Subsetting and summarising</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->


<!--- ---------------------------------------------------------------------- -->
<!--- 2.1 The dplyr package -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2.1">The dplyr package</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

The dplyr package contains some very helpful functions to manipulate data, including subsetting and summarising:  
<br>
``` {r, echo=FALSE}
explore <- data.frame("function name"=c("``select()``", "``filter()``", "``summarise()``", "``group_by()``", "``mutate()``"),
                      "description"=c("subsets columns based on their names",
                                      "subsets rows based on their values",
                                      "calculates summary statistics",
                                      "groups variables for summarising",
                                      "adds new columns that are functions of existing variables"))
kable_styling(kable(explore, col.names=c("function name", "description")), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2.2 Subsetting -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2.2">Subsetting</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4>``select()``</h4>
<ul>
<li>The `` select()`` function is used to select columns of data by their column name.</li>  
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
select(raw_data, date, region, x, y, species, density)
```

``` {r, echo=FALSE}
select_1 <- data.frame(select(raw_data, date, region, x, y, species, density))
kable_styling(kable(select_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```

<br><br>
<h4>``filter()``</h4>
<ul>
<li>The `` filter()`` function is used to select rows based on their values.</li>
<li>To select the observations you want, it is useful to know some comparison operators:</li>
</ul>
``` {r, echo=FALSE}
explore <- data.frame("operator"=c(">", ">=", "!=", "==", "&", 
                                   "|", "!", "``%in% c(...)``"),
                      "description"=c("greater than", "greater than or equal to", "not equal",
                                      "equal", "and", "or", "not", "one in a list of elements"))
kable_styling(kable(explore, escape=FALSE), 
              font_size=14, full_width=FALSE, position="left", 
              bootstrap_options=c("bordered", "condensed"), protect_latex = TRUE)
```

<em>E.g.</em>
```{r, eval=FALSE}
filter(raw_data, region == "Mara" & age <= 5)
```

``` {r, echo=FALSE}
filter_1 <- data.frame(filter(raw_data, region == "Mara" & age <= 5))
kable_styling(kable(filter_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2.3 Summarising -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2.3">Summarising</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4> ``summarise()``</h4>
<ul>
<li>The `` summarise()`` function is used to calculate summaries of variables in your data.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
summarise(raw_data, total_cases = n(), mean_age = mean(age))
```

``` {r, echo=FALSE}
summarise_1 <- data.frame(summarise(raw_data, total_cases = n(), mean_age = mean(age)))
kable_styling(kable(summarise_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```

<br>
<h4> ``group_by()``</h4>
<ul>
<li>The `` group_by `` function is used to group variables before summarising.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
grouped_data <- group_by(raw_data, species)
summarise(grouped_data, total_cases = n(), mean_age = mean(age))
```

``` {r, echo=FALSE}
grouped_1 <- group_by(raw_data, species)
summarise_2 <- data.frame(summarise(grouped_1, total_cases = n(), mean_age = mean(age)))
kable_styling(kable(summarise_2), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed"))
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2.4 Adding new variables -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2.4">Adding new variables</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<h4>``mutate()``</h4>
<ul>
<li>The ``mutate()`` function adds new columns that are functions of existing variables.</li>
</ul>
<em>E.g.</em>
```{r, eval=FALSE}
raw_data <- mutate(raw_data, animal_type = ifelse(species == "human", "human", "animal"))
```

``` {r, echo=FALSE}
mutate_1 <- mutate(raw_data, animal_type = ifelse(species == "human", "human", "animal"))
kable_styling(kable(mutate_1), 
              font_size=14, full_width=FALSE, position="left", bootstrap_options=c("bordered", "condensed")) %>%
  scroll_box(height="200px")
```
<br>

<!--- ---------------------------------------------------------------------- -->
<!--- 2.5 The ``%`` (pipe) function -->

<div class="row"> <!--- Setup grid for layout -->
<div class="col-md-12"> <!--- Begin column set -->
<div class="panel panel-section"> <!--- Begin panel -->
<h3 id="2.5">The ``%`` (pipe) function</h3>
</div> <!--- End panel -->
</div> <!--- End column set -->
</div> <!--- End grid -->

<ul>
<li>dplyr functions can be used individually, or can be 'stacked' using the ``%`` (pipe) function:
<br>

```{r, eval=FALSE}
df_filtered <- filter(raw_data, species != "human") 
df_grouped <- group_by(df_filtered, animal_type)
df_summarised <- summarise(df_grouped, n())

df_summarised <-  raw_data %>% 
  filter(species != "human") %>% 
  group_by(animal_type) %>% 
  summarise(n())
```
</li>
<li>Using the pipe function makes the computing more efficient as it passes the output of one function directly into the next.</li>
<li>This reduces the number of objects you save to the global environment.</li>
</ul>
<br>
